SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2012-12-27 paper/paper.tex
paper: updates on extrapolation.

@@

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.  Indeed, because of the regular structure
of \ttp{shrink} implementations, the principal motivation for one
of Haskell's first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

@@

To present the algorithm, consider the \ttp{SubTypes} type class below for which
QuickCheck's \ttp{Arbitrary} class is a super-class.\footnote{This presentation
  deviates from the implementation; for example, we elide details regarding
  error handling.}
%
\begin{code}
type Idx = Int

class Arbitrary a => SubTypes a where
  size    :: a -> Idx
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
\end{code}
%
\noindent
\ttp{SubTypes} has three methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value, and
\ttp{replace} replaces a sub-value at a particular focus.  If the index is
out-of-bounds (too large or negative), \ttp{index} return \ttp{Nothing}.  The
method \ttp{replace} returns the original value for an out-of-bounds index or an
incorrectly-typed replacement value.  The \ttp{SubVal} type is an existential
data type, used as a generic container for sub-values from a counterexample.
%
\begin{code}
data SubVal =
  forall a. SubTypes a => SubVal a
\end{code}
%
\noindent
For example, consider a binary tree type
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes}; in the following, we show
the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{code}
reduce :: SubTypes a
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

newVals :: SubVal -> IO [SubVal]
newVals (SubVal a) =
  sequence (replicate max s)
  where
  s = sizedArbitrary a >>= return . SubVal

@@

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{maxTries} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

@@

  test' v idx = do
    vs <- newVals v
    case test cex idx vs prop of
      Nothing   -> reduce' (idx+1)
      Just cex' -> reduce0 cex' prop

@@

\subsection{Universal Sub-Value Generalization}
An example best explains the first kind of quantified formulas we generate.
Compare the following formula to the counterexample in Figure~\ref{fig:div-cex}:
%
\begin{code}
\ttp{forall x, y . Div (Div x (Div (C 2) (C 13))) y}
\end{code}
%
\noindent
Here, the formula claims that for all well-typed values \ttp{x} and \ttp{y}, the
property fails.  Not only does the generalization assist the programmer's
insight, but it reduces the sheer size of the counterexample, and it can do so significantly.

\begin{figure}
  \begin{code}
extrapolate :: SubTypes a
       => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- extrapVals v
        if allFail cex idx vs prop
          then extrapolate' (idx+1) (idx:idxs)
          else extrapolate' (idx+1) idxs

extrapVals :: SubVal -> IO [SubVal]
extrapVals (SubVal a) =
  sequence (replicate maxExtrap s)
  where
  s = unsizedArbitrary a >>= return . SubVal

allFail :: SubTypes a
     => a -> Idx -> [SubVal]
     -> (a -> Property) -> Bool
allFail cex idx vs prop =
  (not (null res)) && and res
  where
  res  = catMaybes (map (fail prop) vals)
  vals = map (replace cex idx) vs
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

We call the kind of formula just shown \emph{universal sub-value generalization}
and is implemented in SmartCheck.  The algorithm is shown in
Figure~\ref{fig:universal}.  The algorithm is similar to the reduction algorithm
in Figure~\ref{fig:reduction} (and in the implementation, the algorithms are
generalized and combined).  The function \ttp{extrapolate} takes a
counterexample \ttp{cex} and a property.  It returns a list of indexes to be
universally quantified.  This algorithm also performs a breadth-first search
over the sub-values of the counterexample.  For each sub-value, the algorithm
generates new sub-values, replaces them in \ttp{cex}, and tests them.  If every
new counterexample non-trivially fails (i.e., does not just fail the
precondition) the property \ttp{prop}, then we extrapolate, claiming that for
\emph{any} new value replacing that sub-value in \ttp{cex}, the property will
fail.

The function \ttp{extrapVals} generates the new sub-values, much like
\ttp{newVal} in the reduction algorithm.  Here, \ttp{maxExtrap} well-typed
values are generated, and \ttp{unsizedArbitrary} is the analogue of
\ttp{sizedArbitrary}, but does not limit the size of values generated.

In the definition of \ttp{allFail},
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is the roughly the dual of \ttp{pass} in the reduction algorithm: it returns
\ttp{Just True} if a value satisfies the property and passes the property's
precondition \ttp{pre}, \ttp{Just False} if the value fails the property, and
\ttp{Nothing} if the value fails \ttp{pre}.

Thus,

@@

>>>>

\todo{note constants are user-defined arguments}

@@

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.\footnote{All examples and algorithms in
  this paper are in Haskell~\cite{haskell98}}.  Indeed, because of the regular
structure of \ttp{shrink} implementations, the principal motivation for one of
Haskell's first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

@@

To present the algorithm, consider the \ttp{SubTypes} type class below for which
QuickCheck's \ttp{Arbitrary} class is a super-class.\footnote{This presentation
  deviates from the implementation; for example, we elide details regarding
  error handling.}
%
\begin{code}
type Idx = Int
type Size = Int

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
\end{code}
%
\noindent
\ttp{SubTypes} has three methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value, and
\ttp{replace} replaces a sub-value at a particular focus.  If the index is
out-of-bounds (too large or negative), \ttp{index} return \ttp{Nothing}.  The
method \ttp{replace} returns the original value for an out-of-bounds index or an
incorrectly-typed replacement value.  The \ttp{SubVal} type is an existential
data type, used as a generic container for sub-values from a counterexample.
%
\begin{code}
data SubVal =
  forall a. SubTypes a => SubVal a
\end{code}
%
\noindent
For example, consider a binary tree type
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
%
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3}
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes}; in the following, we show
the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

@@

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{code}
reduce :: SubTypes a
       => a -> (a -> Property) -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of
      Nothing -> return cex
      Just v  -> do
        vs <- newVals v (getSize v) scMaxFailure
        case test cex idx vs prop of
          Nothing   -> reduce' (idx+1)
          Just cex' -> reduce cex' prop

getSize :: SubVal -> Size
getSize (SubVal a) = size a

newVals :: SubVal -> Size -> Int -> IO [SubVal]
newVals (SubVal a) sz tries =
  sequence (replicate tries s)
  where
  s  =     sizedArbitrary sz a
       >>= return . SubVal

@@

\ttp{sizedArbitrary} returns a value inside the \ttp{IO} monad since it depends
on a random seed for generation (we could pass in the generator seed, but for
simplicity, we assume one is generated locally).  The constant \ttp{scMaxFailure} is
the maximum number of tries to discover a new counterexample by replacing
\ttp{v} in \ttp{cex} and testing it.  The result of \ttp{pass prop cex'} for
%
\begin{code}
pass :: (a -> Property) -> a -> Bool
\end{code}
%
\noindent
holds if \ttp{cex'} satisfies the property \ttp{prop}.  If no failure is found,
we move to the next sub-value of \ttp{cex} and continue.  However, if a new
smaller counterexample \ttp{cex'} is found, we start a new breadth-first
traversal of \ttp{cex'}, attempting to shrink it further.

@@

  test' v idx = do
    vs <- newVals v scMaxFailure
    case test cex idx vs prop of
      Nothing   -> reduce' (idx+1)
      Just cex' -> reduce0 cex' prop

@@

\subsection{Universal Sub-Value Generalization}
An example best explains the first kind of quantified formulas we generate.
Compare the following formula to the counterexample in Figure~\ref{fig:div-cex}:
%
\begin{code}
\ttp{forall x, y . Div (Div x (Div (C 2) (C 13))) y}
\end{code}
%
\noindent
Here, the formula claims that for all well-typed values \ttp{x} and \ttp{y}, the
property fails.  Not only does the generalization assist the programmer's
insight, but it reduces the sheer size of the counterexample, and it can do so
significantly.  We call the kind of formula just shown \emph{universal sub-value
  generalization} and is implemented in SmartCheck.

An extrapolation algorithm performs universal sub-value generalization.  The
basic idea is as follows: for a counterexample \ttp{cex} and a property
\ttp{prop}, a breadth-first search over the sub-values of the \ttp{cex} is
performed.  For each sub-value, the algorithm generates new sub-values, replaces
them in \ttp{cex} to create a list of new potential counterexamples.  If enough
new counterexamples non-trivially fails \ttp{prop} (i.e., does not just fail the
precondition), and non satisfy \ttp{prop}, then we extrapolate, claiming that
for \emph{any} new value replacing that sub-value in \ttp{cex}, the property
will fail.

\begin{figure}
  \begin{code}
extrapolate :: SubTypes a
       => a -> (a -> Property) -> IO [Idx]
extrapolate cex prop = extrapolate' 1 []
  where
  extrapolate' idx idxs =
    case index cex idx of
      Nothing -> return idxs
      Just v  -> do
        vs <- newVals v scMaxSize scMaxSuccess
        if allFail cex idx vs prop
          then extrapolate' (idx+1) (idx:idxs)
          else extrapolate' (idx+1) idxs

allFail :: SubTypes a
     => a -> Idx -> [SubVal]
     -> (a -> Property) -> Bool
allFail cex idx vs prop =
  length res >= scMinExtrap && and res
  where
  res  = catMaybes (map go vs)
  go   = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  New sub-values are generated by \ttp{newVals}, shown
in Figure~\ref{fig:reduction}.  Here, we bound the size of values to generate by
the constant \ttp{scMaxSize}, which is independent of the size of the particular
sub-value.

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if enough
new values pass the property's precondition and fail the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is the roughly the dual of \ttp{pass} in the reduction algorithm: \ttp{fail prop
  cex} returns \ttp{Just True} if \ttp{cex} passes \ttp{prop}'s precondition but
fails the property; \ttp{Just False} if \ttp{cex} non-trivially satisfies
\ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails \ttp{prop}'s precondition.  The
constant \ttp{scMinExtrap} is the minimum number of \ttp{Just False} results
required from \ttp{fail} to extrapolate from failed tests to a universal claim.

@@

\todo{check all option names}
